A Constructive Ordered Predicate Logic for Category Theory

### 2022-10-13 paper/paper.tex
work on substitutions in the appendix

Profunctors are very useful for formalizing category theory, but an
additional reason we make them a basic concept of \vett{} is that they
allow us to give a \emph{universal property} for the type of ``morphisms
in a category ${\cat C}$''.  This is analogous to how the $J$
elimination rule for the identity type in Martin-L\"of type theory gives
a universal property for morphisms in a groupoid (the special case of a
category where all morphisms are
invertible)~\cite{hofmannstreicher,awodeywarren,voevodsky}. The reason
profunctors are useful for this purpose is that, for any category $\cat
C$, $\text{Hom}_{\cat C} : \cat C \pto \cat C$ is a profunctor. On
preorders this is just the preorder's ordering relation
itself. Moreover, the hom profunctor is the unit for a composition of
profunctors $R \odot S$ which is defined as a \emph{co-end}.
%% The
%% heteromorphisms of $(R \odot S)(c,e)$ are triples $(d \in \cat D, f \in
%% R(c,d), g \in S(d,e))$ modulo the equivalence relation generated by
%% $(d,f,g) ~ (d', f', g')$ if there exists $h \in D(d,d')$ such that
%% $h\circ f = f'$ and $g'\circ h = g$.
The composition of profunctors is a generalization of the composition of
relations, and just as the equality relation is the identity for the
composition of relations, the hom profunctor is the identity for this
composition.  The unit law for the hom profunctor can be seen as a
``morphism induction'' principle, analogous to the ``path induction''
used in homotopy type theory (though in this paper we consider only
ordinary 1-dimensional categories, not higher generalizations).

@@

    \inferrule*[right=Cat Subst formation]
    {\Gamma \vdash \Psi \isavectx \and \Gamma \vdash \Phi \isavectx}
    {\Gamma \vdash \phi : \Psi \to \Phi}

  \end{mathpar}
  \caption{Substitution}

\end{figure}

\begin{figure}
  \begin{mathpar}
    \inferrule{}{\cdot \isadtctx}

@@

    \inferrule
    {\Psi = \Psi_1 \jnctx \Psi_2\and
      \Gamma\pipe\Psi_1 \vdash \phi : \Phi\and
      \Gamma \pipe \Psi_2 \vdash t : R[d^+\phi;b/\beta] \and
      \Gamma \vdash \phi_2 : \Psi_2 \to \Phi_2
      }
    {\Gamma\pipe\Psi \vdash \phi,t/x,b/\beta :: (\Phi,x:R,\beta:\cat D)}
  \end{mathpar}
  \begin{align*}
    \underline{\alpha:C} &= \alpha:C\\
    \underline{\Phi,x:R,\Psi} &= d^-\Phi; d^+{\Psi}
  \end{align*}
  \begin{align*}
    d^{\pm}(\alpha:\cat C) &= \alpha : \cat C\\
    d^{-}(\Phi,x:R,\Psi) &= d^-\Phi\\
    d^{+}(\Phi,x:R,\Psi) &= d^-\Psi
  \end{align*}
  \begin{align*}
    d^{\pm}(a/\alpha) &= a/\alpha\\
    d^{-}(\phi,t/x,\psi) &= d^-\phi\\
    d^{+}(\phi,t/x,\psi) &= d^+\psi\\
  \end{align*}
  \caption{Basic Rules of Judgment Theory}
  \label{fig:contexts}
\end{figure}

@@

    Composition $(a,t,b) \circ (a',t',b') = (a[a'],t[t'],b[b'])$ is
    given by component-wise single-place substitution with the triple
    of all variables the identity.
  \item The inclusion functor $i^\Gamma : C_d \to C_1$ is defined on
    types as the singleton context $i(R) = \alpha:\cat
    C,x:R,\beta:\cat D$ and on morphisms as the singleton
    substitution.
  \item The restriction operation $R \circ (a,b)$ is defined as the
    substitution $R[a;b]$.
  \item The remaining universal properties are satisfied using the
    analogous type theoretic construction.
  \end{enumerate}
\end{construction}

The associativity, unit and interchange laws of a category are proven
in the \cref{appendix}.

\begin{theorem}{(Sketch) Soundness/Initiality}
  Given any HED $\mathcal V$, there is a unique homomorphism of HEDs $\sem\cdot$
  from $\textrm{Syn}(\textrm{SCT})$ to $\mathcal V$.
\end{theorem}
\begin{proof}
  Roughly the converse of the completeness theorem. For instance,
  profunctor contexts are interpreted using horizontal identity and
  composition as well as the $i$ functor. Preservation of the
  restriction operation requires that restriction commutes with the
  display connectives.
\end{proof}

Details of the interpretation of profunctor contexts and substitutions
are given in the appendix\cref{appendix}.

\subsection{Example Models}

We now describe how to interpret the syntax in any virtual equipment
(that can model the connectives). First, from any virtual equipment we
can construct a DCD:

\begin{definition}
  Given a virtual equipment $\mathcal V$, define a DCD by taking the
  horizontal morphisms of $\mathcal V$ as the display morphisms and
  the horizontal morphisms as the ``composable strings'' of horizontal
  morphisms. Then we define $C_\vdash$ using the notion of
  ``multi-2-cell''.
\end{definition}

Next, we can upgrade any DCD in $\Set$ to a DCD internal to
$\Set^{\Set^o}$ by taking the presheaf of objects, vertical arrows, etc as
families of objects, vertical arrows etc indexed by the indexing set
$\Gamma$ as follows:
\begin{definition}[Powers of Virtual Equipments]
  Let $\mathcal V$ be a virtual equipment and let $X$ be a set.
  Define a virtual equipment, the ``power'' $\mathcal V^X$ as follows.

  \begin{enumerate}
  \item Objects are functions $X \to \mathcal V_0$.
  \item Vertical morphisms $\mathcal V^X_0(A,B)$ are functions
    $\Pi_{x\in X} V_0(A x,B x)$. With pointwise composition, identity.
  \item Horizontal morphisms $\mathcal V^X_1$ are functions $\Gamma
    \to \mathcal V_1$.
  \item A 2-cell from $\Phi$ to $R$ is a family $\Pi_{x\in X} \mathcal
    V^X_h(\Phi x,R x)$.
  \end{enumerate}
  All operations are lifted pointwise.
\end{definition}

This defines the ``natural'' interpretation of SCT in a virtual
equipment once we see that the powering construction is functorial in
$\Gamma$.
\begin{definition}[Internal Logic]
  Given any virtual equipment $\mathcal V$, we can construct the
  ``internal logic'' by indexing $\mathcal V$ by $\Set$:

  \begin{enumerate}
  \item The indexing cateogry is $\Set$, which is a category with families.
  \item The equipment is defined as $\mathcal U\Gamma = \mathcal V^\Gamma$.
  \item Restriction by a function $\gamma : \Delta \to \Gamma$ is
    given pointwise.
  \item The representing object for objects is $\textrm{Ob}_\Gamma =
    \textrm{Ob}(\mathcal V)^\Gamma$ and similarly for the other
    universe objects.
  \end{enumerate}
\end{definition}

\section{Completeness}

@@

    \inferrule{\Gamma \vdash \xi : Z \to \Xi\and \Gamma \pipe \Xi \vdash R \isaSet}{\Gamma \pipe Z \vdash R[\xi] \isaSet}

    \inferrule{\Gamma \vdash \phi : \Psi \to \Phi\and \Gamma \pipe \Phi \vdash M : R}{\Gamma \pipe \Psi \vdash M[\phi] : R[\underline \phi]}
  \end{mathpar}
  \begin{align*}
    (\Phi \jnctx \Psi) \jnctx \Sigma &= \Phi \jnctx (\Psi \jnctx \Sigma)\\
    \Phi \jnctx \alpha:C &= \Phi\\
    \alpha:C \jnctx \Psi &= \Psi\\\\
    d^\pm(A/\alpha) &= A/\alpha\\
    d^+(\phi_1,M/x,\phi_2) &= d^+ \phi_2\\
    d^-(\phi_1,M/x,\phi_2) &= d^- \phi_1\\\\
    \underline \cdot &= \cdot\\
    \underline {A/\alpha} &= A/\alpha\\
    \underline {\phi_1,M/x,\phi_2} &= d^-\phi_1;d^+\phi_2
  \end{align*}
  \caption{Admissible Rules}
\end{figure}

>>>>

Profunctors are very useful for formalizing category theory, but an
additional reason we make them a basic concept of \vett{} is that they
allow us to give a \emph{universal property} for the type of ``morphisms
in a category ${\cat C}$''.  This is analogous to how the $J$
elimination rule for the identity type in Martin-L\"of type theory gives
a universal property for morphisms in a groupoid (the special case of a
category where all morphisms are
invertible)~\cite{hofmann98groupoid,awodeywarren09identity,voevodsky11wollic}. The reason
profunctors are useful for this purpose is that, for any category $\cat
C$, $\text{Hom}_{\cat C} : \cat C \pto \cat C$ is a profunctor. On
preorders this is just the preorder's ordering relation
itself. Moreover, the hom profunctor is the unit for a composition of
profunctors $R \odot S$ which is defined as a \emph{co-end}.
%% The
%% heteromorphisms of $(R \odot S)(c,e)$ are triples $(d \in \cat D, f \in
%% R(c,d), g \in S(d,e))$ modulo the equivalence relation generated by
%% $(d,f,g) ~ (d', f', g')$ if there exists $h \in D(d,d')$ such that
%% $h\circ f = f'$ and $g'\circ h = g$.
The composition of profunctors is a generalization of the composition of
relations, and just as the equality relation is the identity for the
composition of relations, the hom profunctor is the identity for this
composition.  The unit law for the hom profunctor can be seen as a
``morphism induction'' principle, analogous to the ``path induction''
used in homotopy type theory (though in this paper we consider only
ordinary 1-dimensional categories, not higher generalizations).

@@

    \inferrule*[right=Cat Subst formation]
    {\Gamma \vdash \Psi \isavectx \and \Gamma \vdash \Phi \isavectx}
    {\Gamma \vdash \phi : \Psi \to \Phi}
  \end{mathpar}
  \begin{mathpar}
    \inferrule{}{\cdot \isadtctx}

@@

    \inferrule
    {\Psi = \Psi_1 \jnctx \Psi_2\and
      \Gamma\pipe\Psi_1 \vdash \phi : \Phi\and
      \Gamma \pipe \Psi_2 \vdash t : R[d^+\phi;b/\beta] \and
      \Gamma \vdash \phi_2 : \Psi_2 \to \Phi_2
      }
    {\Gamma\pipe\Psi \vdash \phi,t/x,b/\beta :: (\Phi,x:R,\beta:\cat D)}
  \end{mathpar}

  \caption{Substitution}
\end{figure}

\begin{figure}
  \begin{align*}
    \underline{\alpha:C} &= \alpha:C\\
    \underline{\Phi,x:R,\Psi} &= d^-\Phi; d^+{\Psi}
  \end{align*}
  \begin{align*}
    d^{\pm}(\alpha:\cat C) &= \alpha : \cat C\\
    d^{-}(\Phi,x:R,\Psi) &= d^-\Phi\\
    d^{+}(\Phi,x:R,\Psi) &= d^-\Psi
  \end{align*}
  \begin{align*}
    d^{\pm}(a/\alpha) &= a/\alpha\\
    d^{-}(\phi,t/x,\psi) &= d^-\phi\\
    d^{+}(\phi,t/x,\psi) &= d^+\psi\\
  \end{align*}
  \caption{Operations on Contexts}
\end{figure}

In Figure~\ref{fig:admissible}, we describe the admissible typing
rules for the actions of substitution.

\begin{definition}[Substitution Actions]
  For any $\Gamma \pipe \alpha: \cat C \vdash a : \cat D$ and $\Gamma
  \pipe \beta:\cat D \vdash b \cat E$, we define $b[a/\beta]$ by
  recursion on $b$:

  \begin{align*}
    \beta[a/\beta] &= a\\
    (M\,b)[a/\beta] &= M\,(b[a/\beta])\\
    (\lambda \beta)    
    (\pi_\pm b)[a/\beta] &= \pi_\pm b[a/\beta]\\
    (b_-,b_+,s)[a/\beta] &= (b_-[a/\beta],b_+[a/\beta], s[a/\beta])\\
  \end{align*}

  Simultaneously by recursion on $s$ we define $s[\phi]$
  \begin{align*}
    x[a/\alpha,t,b/\beta] &= t\\
    \pendappXtoY M b[a/\alpha] &= \pendappXtoY M {b[a/\alpha]}\\
    {\punitelimkontZatABC {\alpha. t}{b_1} s {b_2}}[\phi] &=
    {\punitelimkontZatABC {\alpha. t}{b_1[d^-\phi]} {s[\phi]} {b_2[d^+\phi]}}\\
    (\punitrefl b)[a/\alpha] &= \punitrefl {b[a/\alpha]}\\
    {\tensorelimWkontZ s {x.\beta.y.r}[\phi_l\jnctx\phi_m\jnctx\phi_r]} &=
    \tensorelimWkontZ {s[\phi_m]} {x.\beta.y. r[\phi_l,x/x,\beta/\beta \jnctx \beta/\beta,y/y,\phi_r]}\\
    {\tensorintroatXwithYandZ b s t}[\phi_s \jnctx \phi_t] &= \tensorintroatXwithYandZ {d^+\phi_s} {s[\phi_s]} {t[\phi_t]}\\
    {(\homrappXtoYatZ s t a)}[\phi_f\jnctx \phi_a] &=
    {\homrappXtoYatZ {s[\phi_f]} {t[\phi_a]} {a[d^+\phi_f]}}\\
    {(\homrlambdaXatYdotZ {x}{\alpha} s)}[\phi] &=
    {\homrlambdaXatYdotZ {x}{\alpha} {s[\phi,x/x,\alpha/\alpha]}}\\
    {(\homlappXtoYatZ t s a)}[\phi_a\jnctx \phi_f] &=
    {\homlappXtoYatZ {t[\phi_a]} {s[\phi_f]} {a[d^-\phi_f]}}\\
    {(\homllambdaXatYdotZ {x}{\alpha} s)}[\phi] &=
    {\homllambdaXatYdotZ {x}{\alpha} {s[\alpha/\alpha,x/x,\phi]}}\\
    (\pi_i s)[\phi] &= \pi_i s[\phi]\\
    (s_1,s_2)[\phi] &= (s_1[\phi], s_2[\phi])\\
    ()[\phi] &= ()
  \end{align*}

  And finally, we define $P[\xi]$ by recursion on $P$:
  \begin{align*}
    (M\,a\,b)[\xi] &= M\,(a[d^-\xi])\,(b[d^+\xi])\\
    (\punitinXfromYtoZ {\cat C} a b)[\xi] &= \punitinXfromYtoZ {\cat C} {a[\xi]}{b[\xi]}\\
    (\tensorexistsXwithYandZ \beta P Q)[\xi] &= \tensorexistsXwithYandZ \beta {P[d^-\xi;\beta/\beta]}{P[\beta/\beta;d^+\xi]}\\
    (\homrallXYtoZ \alpha R P)[\xi] &= \homrallXYtoZ \alpha {R[d^+\xi;\alpha/\alpha]} {P[d^-\xi;\alpha/\alpha]}\\
    (\homlallXYtoZ \alpha P Q)[\xi] &= \homlallXYtoZ \alpha {P[\alpha/\alpha;d^-\xi]} {Q[\alpha/\alpha;d^+\xi]}\\
    1[\xi] &= 1\\
    (P_1 \times P_2)[\xi] &= P_1[\xi] \times P_2[\xi]
  \end{align*}
\end{definition}

%% Theorems about substitution?

%% \begin{definition}[Substitution of Objects in Sets]
%%   3
%% \end{definition}
%% \begin{definition}[Substitution of Elements in Elements]
%%   3
%% \end{definition}

    \inferrule*[right=BoundarySubst]{\Gamma \vdash \xi : Z \to \Xi\and \Gamma \pipe \Xi \vdash R \isaSet}{\Gamma \pipe Z \vdash R[\xi] \isaSet}

    \inferrule*[right=TransSubst]{\Gamma \vdash \phi : \Psi \to \Phi\and \Gamma \pipe \Phi \vdash s : R}{\Gamma \pipe \Psi \vdash s[\phi] : R[\underline \phi]}
  \end{mathpar}
  \begin{align*}
    (\Phi \jnctx \Psi) \jnctx \Sigma &= \Phi \jnctx (\Psi \jnctx \Sigma)\\
    \Phi \jnctx \alpha:C &= \Phi\\
    \alpha:C \jnctx \Psi &= \Psi\\\\
    d^\pm(A/\alpha) &= A/\alpha\\
    d^+(\phi_1,M/x,\phi_2) &= d^+ \phi_2\\
    d^-(\phi_1,M/x,\phi_2) &= d^- \phi_1\\\\
    \underline \cdot &= \cdot\\
    \underline {A/\alpha} &= A/\alpha\\
    \underline {\phi_1,M/x,\phi_2} &= d^-\phi_1;d^+\phi_2
  \end{align*}
  \caption{Admissible Rules}
\end{figure}

\begin{figure}

  \caption{Basic Rules of Judgment Theory}
  \label{fig:contexts}
\end{figure}

@@

    Composition $(a,t,b) \circ (a',t',b') = (a[a'],t[t'],b[b'])$ is
    given by component-wise single-place substitution with the triple
    of all variables the identity.
  \item The inclusion functor $i^\Gamma : C_d \to C_1$ is defined on
    types as the singleton context $i(R) = \alpha:\cat
    C,x:R,\beta:\cat D$ and on morphisms as the singleton
    substitution.
  \item The restriction operation $R \circ (a,b)$ is defined as the
    substitution $R[a;b]$.
  \item The remaining universal properties are satisfied using the
    analogous type theoretic construction.
  \end{enumerate}
\end{construction}\section{Completeness}

@@
